---
title: "Do it yourself"
author: "Heather Dye"
date: "October 15, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Software

Excel, Rstudio, and tabula to convert pdfs to csv files.
Tabula: <http://tabula.technology/>

R packages:  

```{r, message=FALSE}
library(geosphere)
library(rgdal)
library(rgeos)
library(ggmap)
library(plyr)
library(tidyr)
library(dplyr)
library(tmap)
```


## Shapefiles

Missouri: <http://geoportal.missouri.edu/geoportal/catalog/search/search.page>

Illinois: <http://www.ilhousedems.com/redistricting/?page_id=554>

The missouri shapefiles are stored in a folder in the working directory called "MO_2016_TIGER_115th_Congressional_Districts_shp ". One potential problem is not setting the projection co-ordinates. Check that the shape file's co-ordinates are being correctly projected. 
```{r shapes, message=FALSE}
mo<-readOGR(dsn="MO_2016_TIGER_115th_Congressional_Districts_shp", layer ="MO_2016_TIGER_115th_Congressional_Districts_shp")

plot(mo)
proj4string(mo)
```
This is where you check that the shapefile is equipped with a method of projection (Coordinate Reference System), and fix it if needed. I wound up switching between two methods to do the computations
```{r assignproj, message=FALSE}
mo<-spTransform(mo, CRS=CRS("+proj=merc +ellps=GRS80 +units=us-mi"))

proj4string(mo)
```


The information included in the data files includes several data fields. We can strip off an individual district and examine it. 
```{r shapes2}
mo1<-mo[mo@data$CD115FP == "01", ]
head(mo1@data)
sapply(mo1@data, class)
plot(mo1)
View(mo1@data)
```

The following commands are samples of the commands used to add in some of the compactness measures to the data frame. To compute perimeter, I had to convert to a longitude/lattitude. To compute area, I need to a mercator project. The majority of the commands are from rgeos. 

```{r compactness}
mo1latlong<-spTransform(mo1,CRS = CRS("+proj=longlat"))
perimeter(mo1latlong)
gArea(mo1)
gLength(mo1)
gCentroid(mo1)
gConvexHull(mo1)
```

We compute some measures. 
```{r polsbypop, message=FALSE}
mo@data$CD115FP<-as.numeric(mo@data$CD115FP)
mat<-matrix(numeric(), nrow=max(mo@data$CD115FP), ncol=6)
colnames(mat)<-c('cd', 'area', 'hullarea', 'perimeter', 'pols', 'hull')

for(i in array(1:max(mo@data$CD115FP))){mat[[i,1]]<-i
mat[[i,2]]<-gArea(mo[mo@data$CD115FP==i, ])
mat[[i,3]]<-gArea(gConvexHull(mo[mo@data$CD115FP==i, ]))}

molatlong<-spTransform(mo,CRS = CRS("+proj=longlat"))
  
for(i in array(1:max(mo@data$CD115FP))){mat[i,4]<- perimeter(molatlong)[i]
mat[i ,5]<-12*mat[ i, 2]/mat[ i,4]^2
mat[i,6]<-mat[i,2]/mat[i,3]}

print(mat)
```


Attach the data to the shapefiles

```{r attach}
mat<-data.frame(mat)
mo@data<-left_join(mo@data, mat, by=c('CD115FP'='cd'))
mo<-spTransform(mo, CRS=CRS("+proj=longlat"))
````

Plot some shapefiles with comparative information about the measures.

```{r plotshapes}
#qtm(shp=mo, fill="pols", fill.palette ="Blues")
qtm(shp=mo, fill=c("pols", "hull"), fill.palette="Blues", ncol=2)
```

## Adding in census data
To obtain census data about congressional districts. 
<https://www.census.gov/mycd/?st=17>

My file is a csv file titled "MissouriCDcensusRace.csv"
```{r csvfiles}
morace<-read.csv("MissouriCDCensusRace.csv", header=TRUE)
morace<-mutate(morace, perwhite=morace$White/morace$Total.population)
```

Now, I attach the data to my shapefile
```{r shapes3}
mo@data<-left_join(mo@data,morace, by=c('CD115FP'='X'))
qtm(shp=mo, fill=c("hull", "perwhite"), fill.palette="Blues", ncol=2)
```

```{r regression}
glresults<-lm(mo@data$hull ~ mo@data$perwhite)
summary(glresults)
plot(mo@data$perwhite, mo@data$hull)
abline(glresults)
```

Counties by congressional districts
<https://www.census.gov/geo/maps-data/data/cd_state.html>
County votes
<http://www.cnn.com/election/results/states/missouri#president>

## Voting Results 
```{r votecsvfiles}
movote<-read.csv("MissouriCDCensusVotes.csv", header=TRUE)
mo@data<-left_join(mo@data,movote, by=c('CD115FP'='districts'))
qtm(shp=mo, fill=c("hull", "perRep"), fill.palette="Blues", ncol=2)
```
#regression on voting results
```{r votesregression}
lmvotes=lm(mo@data$perRep ~ mo@data$hull)
plot(mo@data$hull, mo@data$perRep)
abline(lmvotes)
summary(lmvotes)
```
