---
title: "Using R and Excel to analyze gerrymandering"
author: "Heather Dye"
date: "October 15, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Background

This demonstration file was developed at the Geometry of Redistricting Conference in Wisconsin. <https://sites.google.com/site/georedistrictingwisc/>
Special thanks to the organizers!

Background information about gerrymandering and measures of compactness are available in this paper: <https://polmeth.polisci.wisc.edu/Papers/compact.pdf>
The area formula defined in the paper is a variation of Heron's formula.

Background information about spatial mapping in R and the regeos package is available in the following sources.
<https://github.com/Robinlovelace/Creating-maps-in-R>
<https://cran.rstudio.com/web/packages/rgeos/rgeos.pdf>

The goal of this tutorial is to provide enough background so that a person with minimal coding experience in R and a general education course in statistics can study gerrymandering in their own community.

## Software

The software that I used included
Excel, Rstudio, and a program called tabula to convert pdfs to csv files.
Tabula: <http://tabula.technology/>

R packages:  

```{r, message=FALSE}
library(geosphere)
library(rgdal)
library(rgeos)
library(ggmap)
#library(tidyr)
library(dplyr)
library(tmap)
```


## Shapefiles

In this demonstration, I used shapefiles that describe the outline of Missouri's congressional districts as "spatial polygons" with attached data. 

Missouri: <http://geoportal.missouri.edu/geoportal/catalog/search/search.page>


The missouri shapefiles are stored in a folder in the working directory called "MO_2016_TIGER_115th_Congressional_Districts_shp ". One problem that I experienced was that the projection co-ordinates were not correctly set. The projection co-ordinates (CRS) need to be set to mercator or longitude in order to compute area. Check that the shape file's co-ordinates are being correctly projected. 
```{r shapes, message=FALSE}
mo<-readOGR(dsn="MO_2016_TIGER_115th_Congressional_Districts_shp", layer ="MO_2016_TIGER_115th_Congressional_Districts_shp")

plot(mo)
proj4string(mo)
```
This command checks that the shapefile is equipped with a method of projection (Coordinate Reference System). In the next command, I set the CRS. I wound up switching between two methods of projection to do the computations
```{r assignproj, message=FALSE}
mo<-spTransform(mo, CRS=CRS("+proj=merc +ellps=GRS80 +units=us-mi"))

proj4string(mo)
```


The information included in the data files includes several data fields. We can strip off an individual district and examine it using commands from dplyr. In Rstudio, the command
View(mo1) will set up a data table for you to inspect. 
```{r shapes2}
mo1<-mo[mo@data$CD115FP == "01", ]
head(mo1@data)
sapply(mo1@data, class)
plot(mo1)
View(mo1@data)
```

The following commands are samples of the commands that I used to to compute compactness measures. To compute area using gArea, you need to access the invidual entries in the dat frame "mo".  Using rgeos, the spatial commands required a mercator projection. The majority of the commands are from rgeos. To compute perimeter, I had to convert to a longitude/lattitude 

```{r compactness}
mo1latlong<-spTransform(mo1,CRS = CRS("+proj=longlat"))
perimeter(mo1latlong)
gArea(mo1)
gLength(mo1)
gCentroid(mo1)
gConvexHull(mo1)
```

We compute some measure of compactness and add them to the data frame. The column
CD115FP identifies the congressional district as a character. I converted this to a numerical value for ease of reference. The Polsby-Popper measure is
A(D)/P(D)^2. The idea is that a circle would have a score of 1/ 4 Pi:  (Pi r^2)/(2 Pi r)^2. All other shapes would have a lower score. The convex hull measure is a ratio of
A(D)/A(convexhull(D)). Convex objects would have a score of 1. 

Here is a convex hull demonstration.
```{r convexhulldemo}
plot(mo1); plot(gConvexHull(mo1), add=TRUE)
```



```{r polsbypop, message=FALSE}
mo@data$CD115FP<-as.numeric(mo@data$CD115FP)
mat<-matrix(numeric(), nrow=max(mo@data$CD115FP), ncol=6)
colnames(mat)<-c('cd', 'area', 'hullarea', 'perimeter', 'pols', 'hull')

for(i in array(1:max(mo@data$CD115FP))){mat[[i,1]]<-i
mat[[i,2]]<-gArea(mo[mo@data$CD115FP==i, ])
mat[[i,3]]<-gArea(gConvexHull(mo[mo@data$CD115FP==i, ]))}

molatlong<-spTransform(mo,CRS = CRS("+proj=longlat"))
  
for(i in array(1:max(mo@data$CD115FP))){mat[i,4]<- perimeter(molatlong)[i]
mat[i ,5]<-12*mat[ i, 2]/mat[ i,4]^2
mat[i,6]<-mat[i,2]/mat[i,3]}

print(mat)
```

## Attach the measures to the shapefiles
Commands from dplyr are used to attach the data to the shapefiles

```{r attach}
mat<-data.frame(mat)
mo@data<-left_join(mo@data, mat, by=c('CD115FP'='cd'))
mo<-spTransform(mo, CRS=CRS("+proj=longlat"))
```

## Plot the shapefiles 

The next command plots the shapefiles with comparative information about the measures.
Additional commands in the Lovelace tutorial can add in background map tiles. 

```{r plotshapes}
#qtm(shp=mo, fill="pols", fill.palette ="Blues")
qtm(shp=mo, fill=c("pols", "hull"), fill.palette="Purples", ncol=2)
```

## Adding in census data
To obtain census data about congressional districts. 
<https://www.census.gov/mycd/?st=17>
Votes by county for the presidential election.
<http://www.cnn.com/election/results/states/missouri#president>

To obtain counties by congressional districts
<https://www.census.gov/geo/maps-data/data/cd_state.html>
I did a very minimal amount of data cleaning in Excel - mainly computing totals by congressional district. In the case where a county was split between congressional districts, I simply added the county to one of the congressional districts. 
The data is stored in  a csv file titled "MissouriCDcensusData.csv"
```{r csvfiles}
mocensus<-read.csv("MissouriCDCensusData.csv", header=TRUE)
mocensus<-mutate(mocensus, perwhite=mocensus$White/mocensus$Total.population)
```

Now, I attach the data to my shapefile
```{r shapes3}
mo@data<-left_join(mo@data,mocensus, by=c('CD115FP'='districts'))
qtm(shp=mo, fill=c("hull", "perwhite"), fill.palette="Blues", ncol=2)
```



Here are the results for the percent of voters who voted Republican in the 2016 election.

```{r votecsvfiles}
qtm(shp=mo, fill=c("hull", "perRep"), fill.palette="Blues", ncol=2)
```

#Regression on convex hull measure versus race and voting results
In the following two code snippets, we examine if there is a significant relationship between the convex hull measure and another variable.

```{r regression}
glresults<-lm(mo@data$hull ~ mo@data$perwhite)
summary(glresults)
plot(mo@data$perwhite, mo@data$hull)
abline(glresults)
```


#regression on voting results
```{r votesregression}
lmvotes=lm(mo@data$perRep ~ mo@data$hull)
plot(mo@data$hull, mo@data$perRep)
abline(lmvotes)
summary(lmvotes)
```
